---
title: "Selecting marker genes and determining abundances"
output: html_notebook
---

```{r}
library(magrittr)
library(tidyverse)
library(here)
```

## read metadata
```{r}
gt <- read_tsv(here::here("data", "geotraces_formatted_metadata.tsv"))
ta <- read_tsv(here::here("data", "tara_formatted_metadata.tsv"))
ts <- read_tsv(here::here("data", "hot_bats.tsv"))
```

```{r}
samples <- bind_rows(select(gt, sampleID), select(ta, sampleID), select(ts, sampleID)) %>% pull
```

## read markers
```{r}
markers_sim <- read_tsv(here::here("data", "simons_markers_combined.tsv"),
                        col_names=c("sampleID", "group", "COG", "total_reads", "count", "marker_length"),
                        col_types="cccddd")

markers_tar <- read_tsv(here::here("data", "tara_markers_combined.tsv"),
                         col_types="cccddd")

d0 <- bind_rows(markers_sim, markers_tar) %>%
  filter(group=="prochlorococcus") %>% 
  mutate(count = replace_na(count, 0))
```

normalize by marker length
need to length norm because we do within sample comparisons
```{r}
d1 <- d0 %>%
  filter(sampleID %in% samples) %>%
  mutate(RPK = count/marker_length * 1e3)
```

## read TBDT counts
```{r}
tbdt_sim <- read_tsv(here::here("data", "simons_TBDT_combined.tsv"),
                        col_names=c("sampleID", "group", "COG", "total_reads", "count", "marker_length"),
                        col_types="cccddd")

tbdt_tar <- read_tsv(here::here("data", "tara_TBDT_combined.tsv"),
                         col_types="cccddd")

s0 <- bind_rows(tbdt_sim, tbdt_tar) %>%
  filter(group=="prochlorococcus") %>% 
  mutate(count = replace_na(count, 0))
```

length normalize
```{r}
s1 <- s0 %>% 
  mutate(RPK_prochlorococcus_TBDT = count/marker_length * 1e3) %>%
  select(sampleID, RPK_prochlorococcus_TBDT, W=count)
```

# Filter samples

remove samples with low prochlorococcus coverage

median of all length-norm markers (some samples are missing markers!)
get length norm median per COG (account for missing COGs in some SIDs)

compute median of set x, fill set up 'to' length 'with' default 0. Also replace
NA 'with'
```{r}
median_fill_up <- function(x, to=length(x), with=0){
    if(to<length(x)) stop("to needs to be >= length(x)")
    x <- as.vector(na.omit(x))
    x <- c(x, rep(with, to-length(x)))
    median(x)
}
```

```{r}
COG_n <- d0$COG %>% 
  unique %>% 
  length

d1s1 <- d1 %>% 
  group_by(group, sampleID) %>%
  summarize(RPK_COG_median=median_fill_up(RPK, COG_n))
```

drop all samples with less than 100X median coverage per Marker
```{r}
d2 <- d1s1 %>% 
  filter(group=="prochlorococcus" & RPK_COG_median > 100) %>%
  ungroup %>% select(sampleID) %>%
  right_join(d1, .)
```

# Filter bad marker families
q10 and q90 markers but this ignores the fact that COGs are missing from some SIDs (not even 0s)
```{r}
d2s1 <- d2 %>%
    replace_na(list(count=0, RPK=0)) %>%
    group_by(group, sampleID) %>%
    summarize(
        RPK_q10=quantile(RPK, probs=.10),
        RPK_q90=quantile(RPK, probs=.90))
```

```{r}
d3 <- d2 %>% left_join(d2s1) %>%
    filter(RPK > RPK_q10 & RPK < RPK_q90)
```

get COGS only present in all samples
```{r}
COGs_universal <- d3 %>% filter(group=="prochlorococcus") %>%
    group_by(COG) %>%
    summarize(COG_n = n()) %>%
    filter(COG_n == max(COG_n)) %>%
    pull(COG)
```

drop non-universal COGs
```{r}
d4 <- d3 %>% filter(COG %in% COGs_universal)
```

```{r}
d4$COG %>% unique %>% length
```

get quantiles
```{r}
d4s1 <- d4 %>% group_by(sampleID, group) %>%
    summarize(
        M=first(total_reads),
        RPK_total_prochlorococcus_reads=sum(RPK),
        RPK_prochlorococcus_marker_median=median(RPK)) %>%
  select(sampleID, RPK_total_prochlorococcus_reads, RPK_prochlorococcus_marker_median, M)
```

```{r}
s2 <- inner_join(d4s1, s1, by="sampleID") %>%
  select(sampleID, RPK_prochlorococcus_TBDT, RPK_total_prochlorococcus_reads,	
         RPK_prochlorococcus_marker_median, W, M) %>%
  mutate(RA=RPK_prochlorococcus_TBDT/RPK_prochlorococcus_marker_median)
```

Compile global siderophore and Pro abundance with time series averaged

```{r}
s2.global <- s2 %>%
  filter(!(sampleID %in% pull(ts, sampleID)))
```

```{r}
read_tsv(here::here("data", "hot_bats_mean.tsv"))
```

```{r}
s2.ts.avg <- semi_join(s2, ts) %>%
  left_join(ts) %>%
  group_by(section, environment) %>%
  summarize(RPK_prochlorococcus_TBDT=mean(RPK_prochlorococcus_TBDT),
            RPK_total_prochlorococcus_reads=mean(RPK_total_prochlorococcus_reads),
            RPK_prochlorococcus_marker_median=mean(RPK_prochlorococcus_marker_median),
            W=round(mean(W)),
            M=round(mean(M)),
            RA=mean(RA)) %>%
  ungroup() %>%
  mutate(sampleID=case_when(section=="BATS" & environment=="surface" ~ "S0582",
                            section=="BATS" & environment=="dcm" ~ "S0583",
                            section=="BATS" & environment=="mixed.layer" ~ "S0584",
                            section=="HOT" & environment=="surface" ~ "S0528",
                            section=="HOT" & environment=="dcm" ~ "S0529",
                            section=="HOT" & environment=="mixed.layer" ~ "S0530",
                            )) %>%
  relocate(sampleID) %>%
  select(-section, -environment)
```

```{r}
bind_rows(s2.global, s2.ts.avg) %>%
  write_tsv(here::here("output", "siderophore_abundance_global.tsv"))
```

Compile timseries siderophore and Pro abundance 
```{r}
semi_join(s2, ts) %>%
  write_tsv(here::here("output", "siderophore_abundance_hot_bats.tsv"))
```

