---
title: "Tara Oceans Formatting, imputing, and finalizing data"
output:   
  github_document:
    toc: yes
    toc_depth: 2
    fig_width: 7
    fig_height: 5
    dev: png
    keep_html: yes
    html_preview: yes
---

```{r message=FALSE, warning=FALSE}
library(here)
library(tidyverse)
library(readxl)
library(janitor)
library(withr)
library(missRanger)
```

# INTRO
We want to combine environmental, biogeographic, biogeochemical, and hydrographic data into a single dataset that can be loaded with ease. The final product will be a single R object that we can load and use right away with statistical models and visualization approaches or other downstream processing.

# READ DATA
metagenome sequence metadata
```{r message=FALSE, warning=FALSE}
tara.samp <- read_tsv(here::here("data", "tara_ebi_metadata.tsv"))
```

Environmental data from the PANGAEA dataset. IMPORTANT: the ENV-WATERCOLUMN appears to be summary values derived from the ENV_DEPTH_SENSORS product and maybe remote observations (eg MODIS). The table we use primarly is the DEPTH_SENSORS table. Also there is a lot of missing and incomplete nutrient data in the PANGAEA dataset. We will try to fill in using the supplementary data from Sunugawa 2015 and/or by imputing.

All data was downloaded from here: https://doi.pangaea.de/10.1594/PANGAEA.836321
```{r message=FALSE, warning=FALSE}
tara.meta.a <- read_tsv(here::here("data", "tara", "pangea-metadata", "dataset875582", "datasets", "TARA_ENV_DEPTH_SENSORS_red.tsv")) %>%
  janitor::clean_names() %>%
  janitor::remove_empty(which = "cols") %>%
  dplyr::rename(pangaea_id=sample_id, tara_station=station,
         ebi_sample_id=sample_id_ena) %>%
  #mutate(tara_station=ifelse(str_detect(tara_station, "TARA_148"), "TARA_148", tara_station)) %>%
  #filter(tara_station!="TARA_148b") %>%
  left_join(select(tara.samp, sampleID, ebi_sample_id), .) %>% 
  distinct()

#colnames(tara.meta.a)
```

```{r message=FALSE, warning=FALSE}
tara.meta.b <- read_tsv(here::here("data", "tara", "pangea-metadata", "dataset875582", "datasets", "TARA_SAMPLES_CONTEXT_ENV-WATERCOLUMN_2.tab"), 
                     skip = 2572) %>%
  janitor::clean_names() %>%
  janitor::remove_empty(which = "cols") %>%
  dplyr::rename(pangaea_id=sample_id_tara_barcode_number_registered_at, tara_station=station_tara_station_number_registered_at,
         ebi_sample_id=sample_id_ena_sample_accession_number) %>%
  #mutate(tara_station=ifelse(str_detect(tara_station, "TARA_148"), "TARA_148", tara_station)) %>%
  #filter(tara_station!="TARA_148b") %>%
  left_join(select(tara.samp, sampleID, ebi_sample_id), .)

#colnames(tara.meta.b)
```

Other random table?? here: https://doi.pangaea.de/10.1594/PANGAEA.875575
```{r message=FALSE, warning=FALSE}
tara.meta.c <- read_tsv(here::here("data", "tara", "pangea-metadata", "TARA_SAMPLES_CONTEXT_ENV-DEPTH-NUT_20170515.tsv")) %>%
  janitor::clean_names() %>%
  janitor::remove_empty(which = "cols") %>%
  #mutate(tara_station=ifelse(str_detect(tara_station, "TARA_148"), "TARA_148", tara_station)) %>%
  #filter(tara_station!="TARA_148b") %>%
  left_join(select(tara.samp, sampleID, ebi_sample_id), .) %>%
  select(sampleID:depth, matches("q50"))

#colnames(tara.meta.c)
```

Companion table from Sunigawa et al 2015 Science. 
```{r message=FALSE, warning=FALSE}
taraW8 <- read_excel(here::here("data", "tara", "sunigawa_et_al_CompanionTables.xlsx"), sheet = "Table W8") %>%
  janitor::clean_names() %>%
  janitor::remove_empty(which = "cols") %>%
  select(pangaea_id=pangaea_sample_id, 
         mean_depth_m, mean_temperature_deg_c, mean_salinity_psu, mean_oxygen_umol_kg,
         po4_umol_l, si_umol_l, mean_depth_max_fluo_m)
```

# FORMAT DATA
```{r}
tara.meta.red.a <- tara.meta.a %>%
  select(sampleID, pangaea_id, tara_station,
         ebi_sample_id, env_feature,
         temp_deg_c_q50,
         sal_q50,
         sigma_theta_kg_m3_q50,
         chl_a_mg_m3_q50_2)

tara.meta.red.b <- tara.meta.b %>%
  select(sampleID, pangaea_id, tara_station,
         ebi_sample_id,
         chlmax_depth=d_chl_m_m_calculated_from_in_situ_senso,
         no3_mmol_l_at_the_depth_of_the_mixed_lay_1,
         no3_mmol_l_at_the_depth_of_maximum_chlor,
         oxygen_mmol_kg_at_the_depth_of_the_mixed_lay_1,
         oxygen_mmol_kg_at_the_depth_of_maximum_chlor) %>%
  group_by(tara_station) %>%
  mutate(chlmax_depth1=mean(chlmax_depth)) %>%
  ungroup()

tara.meta.red.c <- tara.meta.c %>%
  select(sampleID, pangaea_id, tara_station, ebi_sample_id, 
         no2_q50, no2no3_q50, po4_q50, sio4_q50) %>%
  mutate(no3_q50=no2no3_q50-no2_q50)
```

```{r}
tara.samp.meta <- left_join(tara.samp, tara.meta.red.a) %>%
  left_join(., tara.meta.red.c) %>%
  group_by(tara_station) %>%
  mutate(lat=round(mean(lat), 3), lon=round(mean(lon), 3)) %>%
  ungroup()
```

Final joining

1. chla_norm is every value divided by chla at the SCML layer
2. dcm.layer is if `chla_norm > 0.5*chla_norm`
3. dcm.max is if `chla_norm >= 0.9*chla_norm`
4. hdb_cl is whether the depth of maximum fluorescence falls into the basic depth ranges determined for the SCML clusters derived from the GEOTRACES dataset

```{r}
tara.samp.meta.w8.wide.darwin <- tara.samp.meta %>%
  left_join(., taraW8, by="pangaea_id") %>%
  #left_join(., darwin.joined.final) %>%
  left_join(., select(tara.meta.red.b, sampleID, chlmax_depth1)) %>%
  mutate(temperature_dissolved_deg.c = temp_deg_c_q50,
         salinity_dissolved_PSS.1978 = sal_q50,
         sigmatheta_dissolved_kg.m3 = sigma_theta_kg_m3_q50,
         chl_dissolved_mg.kg = chl_a_mg_m3_q50_2/1000/1.025,
         oxygen_dissolved_umol.kg = mean_oxygen_umol_kg,
         phosphate_dissolved_umol.kg = po4_q50/1.025,
         silicate_dissolved_umol.kg = sio4_q50/1.025,
         nitrite_dissolved_umol.kg = no2_q50/1.025,
         nitrate_dissolved_umol.kg = no2no3_q50/1.025) %>%
  group_by(tara_station, lat, lon) %>%
  mutate(chla_norm=chl_dissolved_mg.kg/max(chl_dissolved_mg.kg), 
         dcm.layer=ifelse(chla_norm>0.5*max(chla_norm), 1, 0),
         dcm.max=ifelse(chla_norm>=0.9*max(chla_norm), 1, 0),
         dcm_depth=ifelse(environment=="dcm", depth, NA),
         hdb_cl=case_when(chlmax_depth1 <= 21.4 ~ 1,
                          chlmax_depth1 > 21.4 & chlmax_depth1 <=62.5 ~ 2,
                          chlmax_depth1 > 62.5 & chlmax_depth1 <=84.9 ~ 3,
                          chlmax_depth1 > 84.9 ~ 4)) %>%
  fill(dcm_depth, .direction = "updown") %>%
  select(sampleID, date, replicate, tara_station, 
         environment, depth, lat, lon, ocean,
         temperature_dissolved_deg.c,
         salinity_dissolved_PSS.1978, 
         sigmatheta_dissolved_kg.m3,
         oxygen_dissolved_umol.kg,
         phosphate_dissolved_umol.kg,
         silicate_dissolved_umol.kg,
         nitrite_dissolved_umol.kg,
         nitrate_dissolved_umol.kg,
         chl_dissolved_mg.kg, chla_norm, dcm_depth, dcm.layer, dcm.max, hdb_cl,
         ebi_run_id, release_version, date_acquired_from_ebi,
         full_sample_name, pangaea_id, env_feature)
```

# OUTPUT
```{r}
write_tsv(tara.samp.meta.w8.wide.darwin, here::here("data", "intermediate_files", "tara_biogeochem_metadata.tsv"))
```
