---
title: "Get data from the MIT Darwin model"
output:   
  github_document:
    toc: yes
    toc_depth: 2
    fig_width: 7
    fig_height: 5
    dev: png
    keep_html: yes
    html_preview: yes
---

```{r message=FALSE, warning=FALSE}
library(here)
library(readr)
library(tidyverse)
library(fuzzyjoin)
library(lubridate)
library(magrittr)
```

Set CMAP API key authorization
mykey = ####
```{r eval=FALSE, include=TRUE}
library(cmap4r)
set_authorization()
```

```{r eval=FALSE, include=TRUE}
local_cat <- get_catalog()
```

# GLOBAL DATASET
Tara Oceans + GEOTRACES

## GET METAGENOME COORDINATES
```{r}
mg <- bind_rows(
  read_tsv(here::here("data", "geotraces", "geotraces_formatted_metadata.tsv"), col_types = c("ccccccDtdddccc")),
  read_tsv(here::here("data", "tara", "tara_formatted_metadata.tsv"), col_types = c("ccccccDtdddccc"))
)

mg.coords <- mg %>%
  select(lat, lon, depth) %>%
  distinct()
```

## DOWNLOAD AND FORMAT DARWIN DATA
Here we use the cmap4r package to get the Darwin model nutrient climatology. We will use this modeled data since it is highly complete and represents more of an "time-averaged" picture rather than the point source observations we have from the GEOTRACES samples.

Table `tblDarwin_Plankton_Climatology`
- Synechococcus (mmol C/L) = `prokaryote_c01_darwin_clim`
- Prochlorococcus (mmol C/L) = `prokaryote_c02_darwin_clim`

Table `tblDarwin_Nutrient_Climatology`
- Dissolved Nitrate (mmol N/L) = `NO3_darwin_clim`
- Dissolved Nitrite (mmol N/L) = `NO2_darwin_clim`
- Dissolved Phosphate (mmol P/L) = `PO4_darwin_clim`
- Dissolved Organic Phosphate (mmol P/L) = `DOP_darwin_clim`
- Dissolved Total Fe (mmol Fe/L) = `FeT_darwin_clim`
- Dissolved Organic Fe (mmol Fe/L) = `DOFe_darwin_clim`
- Particulate Organic Fe (mmol Fe/L) = `POFe_darwin_clim`

### Total dFe climatologial mean
```{r eval=FALSE, include=TRUE}
darwinFe.clim <- get_spacetime(tableName = 'tblDarwin_Nutrient_Climatology',
              varName = 'FeT_darwin_clim',
              dt1='2008-01-01',
              dt2='2008-12-31',
              lat1=-63,
              lat2=50,
              lon1=-180,
              lon2=180,
              depth1=0,
              depth2=300)

darwinFe.clim1 <- darwinFe.clim %>% 
  group_by(lat, lon, depth) %>%
  summarize(value=mean(FeT_darwin_clim, na.rm=T)) %>%
  ungroup() %>%
  rename(FeT_darwin_clim=value)
```

This step just saves you time from having to do the fuzzy_geo_join later
```{r}
darwinFe.clim.joined <- geo_left_join(mg.coords, darwinFe.clim1, 
                        by = c("lat","lon"),
                        distance_col="geojoin.dist.km",
                        max_dist = 65)

darwinFe.clim.joined.final <- darwinFe.clim.joined %>%
  mutate(depth.diff=abs(depth.x-depth.y)) %>%
  group_by(lat.x, lon.x, depth.x) %>%
  filter(geojoin.dist.km==min(geojoin.dist.km)) %>%
  filter(depth.diff==min(depth.diff)) %>%
  slice(1) %>%
  ungroup() 

coord.info <- left_join(rename(darwinFe.clim.joined.final, lat=lat.x, lon=lon.x, depth=depth.x), mg) %>%
  select(sampleID, lat=lat.y, lon=lon.y, depth=depth.y, sample_category) %>%
  filter(sample_category!="timeseries")
  
write_tsv(coord.info, here::here("data", "darwin_clim_mean", "darwin_coords_depth.tsv"))
```

Now save...
```{r}
darwinFe.clim.joined <- left_join(coord.info, darwinFe.clim1) %>%
  select(sampleID, FeT_darwin_clim) %>%
  arrange(sampleID)

write_tsv(darwinFe.clim.joined, here::here("data", "darwin_clim_mean", "dFeT.tsv"))
```

### Nitrate
Read back coordinates if necessary
```{r}
coord.info <- read_tsv(here::here("data", "darwin_clim_mean", "darwin_coords_depth.tsv"))
```

```{r eval=FALSE, include=TRUE}
darwinNO3.clim <- get_spacetime(tableName = 'tblDarwin_Nutrient_Climatology',
              varName = 'NO3_darwin_clim',
              dt1='2008-01-01',
              dt2='2008-12-31',
              lat1=-63,
              lat2=50,
              lon1=-180,
              lon2=180,
              depth1=0,
              depth2=300)

darwinNO3.clim %<>% group_by(lat, lon, depth) %>%
  summarize(value=mean(NO3_darwin_clim, na.rm=T)) %>%
  ungroup() %>%
  mutate(variable="NO3_darwin_clim")

darwinNO3.clim.joined <- left_join(coord.info, darwinNO3.clim) %>%
  select(sampleID, value, variable)

saveRDS(darwinNO3.clim.joined, here::here("combine-impute-finalize", "input", "darwin_clim", "darwinNO3.clim"))
```

### Nitrite
```{r eval=FALSE, include=TRUE}
coord.info <- readRDS(here::here("data", "darwin_clim", "darwin.coords.depth"))

darwinNO2.clim <- get_spacetime(tableName = 'tblDarwin_Nutrient_Climatology',
              varName = 'NO2_darwin_clim',
              dt1='2008-01-01',
              dt2='2008-12-31',
              lat1=-63,
              lat2=50,
              lon1=-180,
              lon2=180,
              depth1=0,
              depth2=500)

darwinNO2.clim %<>% group_by(lat, lon, depth) %>%
  summarize(value=mean(NO2_darwin_clim, na.rm=T)) %>%
  ungroup() %>%
  mutate(variable="NO2_darwin_clim")

darwinNO2.clim.joined <- left_join(coord.info, darwinNO2.clim) %>%
  select(sampleID, value, variable)

write_tsv(darwinNO2.clim.joined, here::here("data", "darwin_clim_mean", "NO2.tsv"))
```

# HOT/BATS TIMESERIES
```{r}
ts <- read_tsv(here::here("data", "hot_bats", "hot_bats_formatted_metadata.tsv"),
               col_types = c("ccccccDtdddccc")) %>%
  mutate(month=month(date))

ts.coords <- select(ts, lat, lon, depth, section) %>% distinct()
```

## HOT
lat = 22.750
lon = -158.000
```{r}
darwinFe.clim.hot <- get_spacetime(tableName = 'tblDarwin_Nutrient_Climatology',
              varName = 'FeT_darwin_clim',
              dt1='2008-01-01',
              dt2='2008-12-31',
              lat1=22,
              lat2=23,
              lon1=-157,
              lon2=-159,
              depth1=0,
              depth2=300)
```

```{r}
darwinFe.clim.hot.joined <- geo_left_join(filter(ts.coords, section=="HOT"), darwinFe.clim.hot, 
                        by = c("lat","lon"),
                        distance_col="geojoin.dist.km",
                        max_dist = 80)

darwinFe.clim.hot.joined.final <- darwinFe.clim.hot.joined %>%
  mutate(depth.diff=abs(depth.x-depth.y)) %>%
  group_by(month, lat.x, lon.x, depth.x) %>%
  filter(geojoin.dist.km==min(geojoin.dist.km)) %>%
  filter(depth.diff==min(depth.diff)) %>%
  slice(1) %>%
  ungroup() 

coord.hot.info <- left_join(ts, rename(darwinFe.clim.hot.joined.final, 
                                   lat=lat.x, lon=lon.x, depth=depth.x)) %>%
  select(sampleID, lat, lon, depth, date, FeT_darwin_clim, section) %>%
  filter(section=="HOT") %>%
  arrange(date)
```

## BATS
lat = 31.667
lon = -64.167	
```{r}
darwinFe.clim.bats <- get_spacetime(tableName = 'tblDarwin_Nutrient_Climatology',
              varName = 'FeT_darwin_clim',
              dt1='2003-01-01',
              dt2='2009-12-31',
              lat1=31,
              lat2=32,
              lon1=-64,
              lon2=-65,
              depth1=0,
              depth2=300)
```

```{r}
darwinFe.clim.bats.joined <- geo_left_join(filter(ts.coords, section=="BATS"), darwinFe.clim.bats, 
                        by = c("lat","lon"),
                        distance_col="geojoin.dist.km",
                        max_dist = 80)

darwinFe.clim.bats.joined.final <- darwinFe.clim.bats.joined %>%
  mutate(depth.diff=abs(depth.x-depth.y)) %>%
  group_by(month, lat.x, lon.x, depth.x) %>%
  filter(geojoin.dist.km==min(geojoin.dist.km)) %>%
  filter(depth.diff==min(depth.diff)) %>%
  slice(1) %>%
  ungroup() 

coord.bats.info <- left_join(ts, rename(darwinFe.clim.bats.joined.final, 
                                   lat=lat.x, lon=lon.x, depth=depth.x)) %>%
  select(sampleID, lat, lon, depth, date, FeT_darwin_clim, section) %>%
  filter(section=="BATS") %>%
  arrange(date)
```

## combine HOT and BATS
```{r}
bind_rows(coord.hot.info,
          coord.bats.info) %>%
  select(sampleID, FeT_darwin_clim) %>%
  write_tsv(here::here("data", "darwin_clim_mean", "dFeT_timeseries.tsv"))
```
