---
title: "HOT BATS Formatting, imputing, and finalizing data"
output:   
  github_document:
    toc: yes
    toc_depth: 2
    fig_width: 7
    fig_height: 5
    dev: png
    keep_html: yes
    html_preview: yes
---

```{r message=FALSE, warning=FALSE}
library(here)
library(readr)
library(tidyverse)
library(fuzzyjoin)
library(lubridate)
library(magrittr)
```

# Read data

## HOT/BATS
```{r eval=FALSE, include=TRUE}
ts <- read_tsv(here::here("data", "hot_bats", "hot_bats_formatted_metadata.tsv"), col_types = c("ccccccDtdddccc")) 
```

## Ecotypes
```{r eval=FALSE, include=TRUE}
ecotypes <- read_tsv(here::here("data", "intermediate_files", "ecotypes_formatted.tsv")) 
```

## Longhursts
```{r eval=FALSE, include=TRUE}
longhurst <- read_tsv(here::here("data", "longhurst_province", "longhurst_codes.tsv"),
                     col_names = c("sampleID", "lat", "lon", "LongCode", "Longtmp")) %>%
  separate(col=Longtmp, into=c("LongWind", "LongDesc"), sep=" - ")
```
## Darwin
```{r}
darwin <- read_tsv(here::here("data", "darwin_clim_mean", "dFeT_timeseries.tsv")) 
```

# Format

Add ecotype abundances, longhurst codes, and darwin climatological average to timeseries
```{r message=FALSE, warning=FALSE}
timeseries.final <- ts %>% 
  mutate(environment=case_when(depth < 150 & depth > 25 ~ "dcm",
                               depth <= 25 ~ "surface",
                               depth > 150 & depth < 300 ~ "mixed.layer",
                               depth > 300 ~ "mesopelagic")) %>%
  mutate(dcm.layer=ifelse(environment=="dcm", 1, 0),
         dcm.max=ifelse(environment=="dcm", 1, 0)) %>%
  select(-depth_category, -replicate, -time, -station_num_BODC, -bottle_num_BODC) %>%
  mutate(date = ymd(date)) %>%
  left_join(., ecotypes) %>%
  left_join(., longhurst) %>%
  left_join(., darwin) %>%
  arrange(section, date, depth) %>%
  mutate(lat=round(lat, 3), lon=round(lon, 3))

write_tsv(timeseries.final, here::here("output", "hot_bats.tsv"))
```

Reduce to a single averaged value for each location/depth. This averaged value will be used in the global dataset. Not appropriate to include correlated/repeated measures with the model we are using espsecially since the repeated measures are very skewed to one location and time range.

```{r}
timeseries.summarized <- timeseries.final %>%
  select(-LongCode, -LongWind, -LongDesc) %>%
  group_by(lat, lon, environment) %>%
  mutate_at(vars(depth,pro_HLI:FeT_darwin_clim), mean, na.rm = TRUE) %>%
  ungroup() %>%
  distinct(lat, lon, environment, .keep_all=TRUE) %>%
  mutate(ocean_category=case_when(section=="BATS" ~ "north atlantic",
                                   section=="HOT" ~ "north pacific")) %>%
  mutate(chla_norm=case_when(environment=="dcm" ~ 1,
                             environment=="mesopelagic" ~ 0))
```

```{r}
write_tsv(timeseries.summarized, here::here("data", "intermediate_files", "hot_bats_mean.tsv"))
```

