---
title: "Combining GEOTRACES and Tara Oceans data, imputing, and finalizing"
output:   
  github_document:
    toc: yes
    toc_depth: 2
    fig_width: 7
    fig_height: 5
    dev: png
    keep_html: yes
    html_preview: yes
---

```{r message=FALSE, warning=FALSE}
library(here)
library(readr)
library(tidyverse)
library(fuzzyjoin)
library(lubridate)
library(magrittr)
library(missRanger)
library(withr)
```

# READ DATA

## ecotypes
```{r}
ecotypes <- read_tsv(here::here("data", "intermediate_files", "ecotypes_formatted.tsv")) 
```

## mean HOT BATS
```{r}
hot_bats_mean <- read_tsv(here::here("data", "intermediate_files", "hot_bats_mean.tsv")) %>%
  mutate(cruise_id=as.character(cruise_id))
```

## Global Darwin Climatological mean Fe
```{r}
darwin.global <- read_tsv(here::here("data", "darwin_clim_mean", "dFeT.tsv")) 
```

## Longhurst provinces
```{r eval=FALSE, include=TRUE}
longhurst <- read_tsv(here::here("data", "longhurst_province", "longhurst_codes.tsv"),
                     col_names = c("sampleID", "lat", "lon", "LongCode", "Longtmp")) %>%
  separate(col=Longtmp, into=c("LongWind", "LongDesc"), sep=" - ")
```

## GEOTRACES CTD data
```{r message=FALSE, warning=FALSE}
GA02.ctd <- read_tsv(here::here("data", "intermediate_files", "GA02_mg_ctd.tsv")) %>% mutate(section="GA02")
GA03.ctd <- read_tsv(here::here("data", "intermediate_files", "GA03_mg_ctd.tsv")) %>% mutate(section="GA03")
GA10.ctd <- read_tsv(here::here("data", "intermediate_files", "GA10_mg_ctd.tsv")) %>% mutate(section="GA10")
GP13.ctd <- read_tsv(here::here("data", "intermediate_files", "GP13_mg_ctd.tsv")) %>% mutate(section="GP13")

ctd <- bind_rows(GA02.ctd, 
                 GA03.ctd, 
                 GA10.ctd, 
                 GP13.ctd)
```

## GEOTRACES bottle data
```{r}
GA02.btl <- read_tsv(here::here("data", "intermediate_files", "GA02_mg_bottle.tsv")) %>% mutate(section="GA02")
GA03.btl <- read_tsv(here::here("data", "intermediate_files", "GA03_mg_bottle.tsv")) %>% mutate(section="GA03")
GA10.btl <- read_tsv(here::here("data", "intermediate_files", "GA10_mg_bottle.tsv")) %>% mutate(section="GA10")
GP13.btl <- read_tsv(here::here("data", "intermediate_files", "GP13_mg_bottle.tsv")) %>% mutate(section="GP13")

btl <- bind_rows(GA02.btl, 
                 GA03.btl, 
                 GA10.btl, 
                 GP13.btl) %>%
  dplyr::select(-station_num_BODC) %>%
  mutate(nitrate_dissolved_umol.kg = ifelse(section=="GA10" & is.na(nitrate_dissolved_umol.kg), (nitrite.nitrate_dissolved_umol.kg-nitrite_dissolved_umol.kg), nitrate_dissolved_umol.kg))
```

## Generic GEOTRACES metadata
```{r}
gt <- read_tsv(here::here("data", "geotraces", "geotraces_formatted_metadata.tsv"))
```
```{r}
gt.no.ts <- left_join(ctd, select(btl, -lat, -lon)) %>% 
  left_join(., select(gt, -bottle_num_BODC, -station_num_BODC, -lat, -lon, -depth, -section))
```

## Tara Oceans
```{r message=FALSE, warning=FALSE}
tara <- read_tsv(here::here("data", "intermediate_files", "tara_biogeochem_metadata.tsv")) %>%
  rename(ocean_category=ocean)
```

## Finish combining all data types
```{r}
combo <- bind_rows(tara, gt.no.ts) %>%
  left_join(., ecotypes) %>%
  left_join(., select(longhurst, -lat, -lon)) %>%
  left_join(., darwin.global) %>%
  bind_rows(., hot_bats_mean)
```

Clean up the ocean category
```{r message=FALSE, warning=FALSE}
combo1 <- combo %>%
  mutate(ocean=case_when(ocean_category == "north atlantic" ~ "atlantic_north",
                         ocean_category == "north pacific"  ~ "pacific_north",
                         ocean_category == "south atlantic" ~ "atlantic_south",
                         ocean_category == "south pacific"  ~ "pacific_south",
                         ocean_category == "mediterranean" ~ "med_red",
                         ocean_category == "red sea" ~ "med_red",
                         ocean_category == "indian" ~ "indian",
                         ocean_category == "southern ocean" ~ "southern_ocean")) %>%
  select(-ocean_category) 
```

Set prochlorococcus ecotype values to zero in all depths deeper than 300 meters. In reality Pro total abundance is so low at those depths that relative ecotype abundances are meaningless
```{r message=FALSE, warning=FALSE}
combo2 <- combo1 %>%
  filter(depth < 301) %>%
  select(-pressure_dissolved_dbar) %>%
  mutate_at(vars(pro_HLI:pro),
             funs(ifelse(depth > 300, 0, .))) %>%
  mutate(LongCode=factor(LongCode),
         LongWind=factor(LongWind),
         ocean=factor(ocean),
         replicate=factor(replicate)) %>%
  mutate(imputed_nitrate=ifelse(is.na(nitrate_dissolved_umol.kg), 1, 0),
         imputed_nitrite=ifelse(is.na(nitrite_dissolved_umol.kg), 1, 0), 
         imputed_oxygen=ifelse(is.na(oxygen_dissolved_umol.kg), 1, 0), 
         imputed_phosphate=ifelse(is.na(phosphate_dissolved_umol.kg), 1, 0),
         imputed_salinity=ifelse(is.na(salinity_dissolved_PSS.1978), 1, 0),
         imputed_silicate=ifelse(is.na(silicate_dissolved_umol.kg), 1, 0),
         imputed_temperature=ifelse(is.na(temperature_dissolved_deg.c), 1, 0)) %>%
  # remove superflous tara metadata
  select(-tara_station, -environment, -date, -hdb_cl, -ebi_run_id, 
         -date_acquired_from_ebi, -full_sample_name, -env_feature, 
         -station_num_BODC, -bottle_num_BODC, -pangaea_id, -geojoin.dist.km,
         -release_version)
```


check for number of missing observations
```{r}
combo2 %>%
  summarise_all(funs(sum(is.na(.))/n()))
```

For cases with sparsely missing data (~25% of observations) we will impute values using a Random Forest with the missRanger package. The missRanger package uses the ranger package to do fast missing value imputation by chained random forest prediction. As such, it serves as an alternative implementation of the 'MissForest' algorithm. In my experience this imputation strategy produces reasonable results in this dataset and allows using much more information in the regression analyses.


```{r}
to.impute <- "dcm.layer + dcm.max + chla_norm + nitrate_dissolved_umol.kg + nitrite_dissolved_umol.kg + oxygen_dissolved_umol.kg + phosphate_dissolved_umol.kg + salinity_dissolved_PSS.1978 + silicate_dissolved_umol.kg + temperature_dissolved_deg.c + pro_LLVIIa + pro_LLVIIb + pro_LLVIIc + pro_LLVIId + pro_LLVIIe + pro_LLVIIf + pel_Ia + pel_Ib + pel_Ic + pel_IIa + pel_IIb + pel_IIIa +  pel_IIIb + pel_IV + pel_V + sar11 + bacteria + archaea + eukaryotes + viruses"

impute.by <- "lat + lon + depth + dcm.layer + dcm.max + chla_norm + ocean + LongCode + LongWind + FeT_darwin_clim + nitrate_dissolved_umol.kg + nitrite_dissolved_umol.kg + oxygen_dissolved_umol.kg + phosphate_dissolved_umol.kg + salinity_dissolved_PSS.1978 + silicate_dissolved_umol.kg + temperature_dissolved_deg.c + pro_HLI + pro_HLII + pro_HLIII_HLIV + pro_LLI + pro_LLII_LLIII + pro_LLIV + pro_LLVII + pro + pro_LLVIIa + pro_LLVIIb + pro_LLVIIc + pro_LLVIId + pro_LLVIIe + pro_LLVIIf + pel_Ia + pel_Ib + pel_Ic + pel_IIa + pel_IIb + pel_IIIa + pel_IIIb + pel_IV + pel_V + sar11 + bacteria + archaea + eukaryotes + viruses"

imputeform <- as.formula(paste(to.impute, "~", impute.by, by=" "))

with_preserve_seed(imp1 <- combo2 %>%
  missRanger(imputeform,
            verbose = 1, 
            num.trees = 10000))
```
# TRACE METAL MEASUREMENTS ONLY

First look at the number of NAs per section. Lots of metals are missing data
```{r}
missing.metals <- btl %>%
  group_by(section) %>%
  select(-sampleID, -lat, -lon, -nitrite.nitrate_dissolved_umol.kg,
         -dissolved.organic.carbon_dissolved_umol.kg, -dissolved.inorganic.carbon_dissolved_umol.kg,
         -salinity_dissolved_PSS.1978, -silicate_dissolved_umol.kg, -phosphate_dissolved_umol.kg, 
         -oxygen_dissolved_umol.kg, -nitrite_dissolved_umol.kg, -nitrate_dissolved_umol.kg,
         -temperature_dissolved_deg.c, -pressure_dissolved_dbar) %>%
  summarise_all(funs(sum(is.na(.))/n())) %>%
  pivot_longer(cols=-section, names_to="variable", values_to="percent_missing")

missing.metals %>% arrange(variable)
```

Metal imputation only GEOTRACES

copper_dissolved_nmol.kg    - GA03, GP13
aluminum_dissolved_nmol.kg  - GA02, GA03
nickel_dissolved_nmol.kg	  - GA02, GA03, GP13
cobalt_dissolved_pmol.kg    - GA02, GA03, GA10
iron_dissolved_nmol.kg      - GA02, GA03, GA10, GP13
lead_dissolved_pmol.kg      - GA02, GA03, GA10, GP13	
manganese_dissolved_nmol.kg	- GA02, GA03, GA10, GP13
zinc_dissolved_nmol.kg      - GA02, GA03, GA10, GP13

## Iron lead, manganese, zinc
```{r}
metals <- btl %>%
  select(sampleID, cobalt_dissolved_pmol.kg, nickel_dissolved_nmol.kg, aluminum_dissolved_nmol.kg,
         copper_dissolved_nmol.kg, iron_dissolved_nmol.kg, lead_dissolved_pmol.kg,
         manganese_dissolved_nmol.kg, zinc_dissolved_nmol.kg, dissolved.inorganic.carbon_dissolved_umol.kg,
         dissolved.organic.carbon_dissolved_umol.kg, molybdenum_dissolved_nmol.kg,
         aluminum_total.particulate_nmol.kg, iron_soluble_nmol.kg, iron_total.particulate_nmol.kg,
         ironII_dissolved_nmol.kg,  L1.copper.bind_dissolved_logK, L1.copper.conc_dissolved_nmol.kg,
         L1.iron.bind_dissolved_logK, L1.iron.conc_dissolved_nmol.kg, L2.iron.bind_dissolved_nmol.kg,
         L2.iron.logk_dissolved_logK, lead_total.dissolvable_pmol.kg, lead_total.particulate_pmol.kg) %>%
  left_join(imp1, ., by="sampleID") #%>%
  mutate(imputed_iron=ifelse(is.na(iron_dissolved_nmol.kg), 1, 0), 
         imputed_lead=ifelse(is.na(lead_dissolved_pmol.kg), 1, 0), 
         imputed_manganese=ifelse(is.na(manganese_dissolved_nmol.kg), 1, 0),
         imputed_zinc=ifelse(is.na(zinc_dissolved_nmol.kg), 1, 0),
         imputed_cobalt=ifelse(is.na(cobalt_dissolved_pmol.kg), 1, 0),
         imputed_nickel=ifelse(is.na(nickel_dissolved_nmol.kg), 1, 0),
         imputed_aluminum=ifelse(is.na(aluminum_dissolved_nmol.kg), 1, 0),
         imputed_copper=ifelse(is.na(copper_dissolved_nmol.kg), 1, 0))
```

```{r message=FALSE, warning=FALSE}
metals.to.impute <- "iron_dissolved_nmol.kg + lead_dissolved_pmol.kg + manganese_dissolved_nmol.kg + zinc_dissolved_nmol.kg + cobalt_dissolved_pmol.kg + nickel_dissolved_nmol.kg + aluminum_dissolved_nmol.kg + copper_dissolved_nmol.kg"

metals.impute.by <- "cobalt_dissolved_pmol.kg + nickel_dissolved_nmol.kg + aluminum_dissolved_nmol.kg + copper_dissolved_nmol.kg + iron_dissolved_nmol.kg + lead_dissolved_pmol.kg + manganese_dissolved_nmol.kg + zinc_dissolved_nmol.kg + dissolved.inorganic.carbon_dissolved_umol.kg + dissolved.organic.carbon_dissolved_umol.kg + molybdenum_dissolved_nmol.kg + aluminum_total.particulate_nmol.kg + iron_soluble_nmol.kg + iron_total.particulate_nmol.kg + ironII_dissolved_nmol.kg +  L1.copper.bind_dissolved_logK + L1.copper.conc_dissolved_nmol.kg + L1.iron.bind_dissolved_logK + L1.iron.conc_dissolved_nmol.kg + L2.iron.bind_dissolved_nmol.kg + L2.iron.logk_dissolved_logK + lead_total.dissolvable_pmol.kg + lead_total.particulate_pmol.kg + lat + lon + depth + dcm.layer + dcm.max + chla_norm + hdb_cl + ocean + LongCode + LongWind + synDarwin_umolC.kg + proDarwin_umolC.kg + nitrateDarwin_dissolved_umol.kg + nitriteDarwin_dissolved_umol.kg + phosphateDarwin_dissolved_umol.kg + DOPDarwin_dissolved_umol.kg + ironDarwin_dissolved_nmol.kg + DOironDarwin_dissolved_nmol.kg + POironDarwin_dissolved_nmol.kg + nitrate_dissolved_umol.kg + nitrite_dissolved_umol.kg + oxygen_dissolved_umol.kg + phosphate_dissolved_umol.kg + salinity_dissolved_PSS.1978 + silicate_dissolved_umol.kg + temperature_dissolved_deg.c + pro_HLI + pro_HLII + pro_HLIII_HLIV + pro_LLI + pro_LLII_LLIII + pro_LLIV + pro_LLVII + pro + pro_LLVIIa + pro_LLVIIb + pro_LLVIIc + pro_LLVIId + pro_LLVIIe + pro_LLVIIf + pel_Ia + pel_Ib + pel_Ic + pel_IIa + pel_IIb + pel_IIIa + pel_IIIb + pel_IV + pel_V + sar11 + bacteria + archaea + eukaryotes + viruses"

metals.imputeform <- as.formula(paste(metals.to.impute, "~", metals.impute.by, by=" "))

with_preserve_seed(metals.imp <- metals %>%
  missRanger(metals.imputeform,
            verbose = 1, 
            num.trees = 10000))
```

```{r}
write_tsv(metals.imp, here::here("output", "geotraces_tara_combined_imputed.tsv"))
```

