---
title: "RELATIONSHIP BETWEEN LLI, NO2, AND DCM_DEPTH"
output: html_notebook
---

```{r}
library(here)
library(tidyverse)
library(ggforce)
library(ggeffects)
library(see)
library(performance)
library(ggeffects)
library(betareg)
library(withr)
library(emmeans)
```

# READ DATA
```{r}
tbdt <- read_tsv(here::here("data", "geotraces_tara_combined_imputed.tsv")) %>%
  group_by(sampleID) %>% 
  slice(1) %>%
  ungroup() %>%
  filter(sampleID != "S0468") %>%
  rename(rpkm_tbdt=RPK_prochlorococcus_TBDT,
         rpkm_pro=RPK_total_prochlorococcus_reads) %>%
  mutate(RA=rpkm_tbdt/rpkm_pro)
```


# RELATIONSHIP BETWEEN LLI, NO2, AND DCM_DEPTH

Relative abundance of the TBDT increases alongside NO2 and LLI but the relationship between DCM_DEPTH is less clear.

```{r}
# cut LLI, NO2, and DCM depth into 3 roughly equally-sized categories
tbdt1 <- tbdt %>%
  mutate(nf=cut_number(nitrite_dissolved_umol.kg, n=3)) %>% #, labels=c("l", "m", "h")
  mutate(dcmf=cut_number(dcm_depth, n=3)) %>% #, labels=c("l", "m", "h")
  mutate(LLIf=cut_number(pro_LLI, n=3)) %>% #, labels=c("l", "m", "h")
  mutate(RA=ifelse(RA==0, rnorm(n(), 3.055513e-07, 1e-7), RA)) %>%
  mutate(pro_LLI_norm=pro_LLI*pro)
```

## EXAMINE DATA
```{r}
ggplot(tbdt1) + 
  geom_jitter(aes(x=nf, y=RA, shape=LLIf, color=pro_LLI), width = 0.25, height=0.0001) + 
  facet_grid(~dcmf) +
  labs(x="NO2 conc factor", y="Rel Abund TBDT") + 
  scale_color_viridis_c(trans="log10") + 
  scale_y_continuous(breaks=c(0.0001, 0.001, 0.002, 0.003), trans="sqrt", limits=c(0, 3.2e-3)) +
  theme_bw() +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank())
```

## MODEL WITH BETA REGRESSION

Idea is to model LLI as constant additive effect while allowing the effect of NO2 to vary along with DCM depth
```{r}
with_preserve_seed(brmod1 <- betareg(
  formula = RA ~ LLIf + nf * dcmf,
  data = tbdt1))

with_preserve_seed(brmod2 <- betareg(
  formula = RA ~ nf * dcmf,
  data = tbdt1))
```

Better to include LLI term
```{r}
compare_performance(brmod1, brmod2, rank = TRUE)
```

# Table S2
```{r}
summary(brmod1)
  
texreg(brmod1)
```

```{r}
brmod1.emm <- emmeans(brmod1, ~ LLIf + nf * dcmf)

contrast(brmod1.emm, "consec", simple = "each", combine = TRUE, adjust = "mvt") %>%
  as.data.frame() %>%
  filter(p.value <= 0.05) #%>%
  xtable::xtable()
```

Kind of like an emmeans-based ANOVA
```{r}
joint_tests(brmod) #%>% xtable::xtable()
```

```{r}
pairs(emmeans(brmod, specs=c("LLIf"), type="re")) #%>% xtable::xtable()
```
```{r}
pairs(emmeans(brmod, specs=c("nf"), type="re"))
```

Get estimated marginal means
```{r}
mydf1 <- ggemmeans(brmod1, terms = c("nf", "LLIf", "dcmf"), interval="confidence") %>%
  rename(dcmf=facet, RA=predicted, nf=x, LLIf=group)
```

# Fig 6
```{r}
fig6 <- ggplot(tbdt1) + 
  geom_jitter(aes(x=nf, y=RA, color=pro_LLI),
              width = 0.25,
              height=0.0001,
              alpha=0.75,
              shape=16) +  
  facet_grid(~dcmf) +
  geom_point(data=mydf1, aes(x=nf, y=RA, shape=LLIf), 
             position = position_dodge(width = 0.9),
             size=2) + 
  geom_linerange(data=mydf1, aes(x=nf, y=RA, group=LLIf, ymin=conf.low, ymax=conf.high), position = position_dodge(width = 0.9)) +
  labs(x="NO2 conc factor", y="Rel Abund TBDT") + 
  scale_color_viridis_c(trans="log10", begin = 0, end=1, option="D") + 
  scale_y_continuous(breaks=c(0.0001, 0.001, 0.002, 0.003), trans="sqrt") +
  theme_bw() +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank())
```

```{r}
ggsave(here::here("figs", "no2_dcm.svg"), plot=fig6,
       device="svg", units="cm", height=7, width=18)
```

